# 6 Analysis of Longitudinal Data

## Task 1: RATS

**Implement the analyses of Chapter 8 of MABS using the RATS data. (0-7 points: 0-4 points for graphs or analysis results + 0-3 points for their interpretations)**

```{r 6_libraries, message=F}
library(dplyr); library(tidyr); library(ggplot2); library(psych); library(kableExtra); library(lme4);
library(GGally); library(corrplot)
options(scipen=999)

RATSL <- read.table("data/RATSL.csv", sep = ",", header = T, row.names = 1)
RATSL <- within(RATSL, {
  Group <- factor(Group)
  ID <- factor(ID)
})
```

<br/>

### The RATS data

The RATS dataset has 16 observations and 5 variables:

<br/>

```{r 6_variables, echo=F, message=F}
vars <- data.frame(
  Variables = colnames(RATSL),
  Information = c(
    "rat id", "treatment group (diet)", "day of measurement from the start of the study",
    "an (approximately) weekly bodyweight in grams", "day of measurement as numeric"
  )
)

kable(vars, caption = "<b>Table 1</b> Variables in the dataset") %>%
  kable_styling(bootstrap_options = c("hover", "striped")) %>%
  column_spec(1, bold = T, italic = T, color = "#008cba") %>% 
  column_spec(2)
```

<br/>

```{r 6_desc}
glimpse(RATSL)
head(RATSL)
tail(RATSL)
summary(RATSL)
describe(RATSL) %>% kable(digits = 2, caption = "<b>Table 2</b> Descriptives") %>%
  kable_styling(bootstrap_options = c("hover", "striped"))
```

<br/>

There are 3 treatment groups, and the measurement of weight has been done during a 9 week period. The mean weight of the rats is about 384 grams and ranges from 225 to 628 grams.

<br/><br/>

### Visualizing the data

```{r 6_fig1, fig.height=5, fig.width=8}
p1 <- ggplot(RATSL, aes(x = Time, y = Weight, group = ID))
p2 <- p1 + geom_line(aes(linetype = Group, color = Group))
p3 <- p2 +  scale_x_continuous(name = "Time (days)", breaks = seq(1, 64, 7)) +
  scale_y_continuous(name = "Weight (grams)")
p4 <- p3 + theme_light() + ggtitle(label = "Figure 1 Repeated measures of weight")
p4
```

<br/>

A quick look at the data shows that the groups have different starting levels but also that the weight gain seems to be stronger in groups 2 and 3 in comparison to group 1.

<br/>

```{r 6_fig2, fig.height=5, fig.width=8}
# Standardising the scores:
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate( stdWeight = (Weight - mean(Weight))/sd(Weight) ) %>%
  ungroup()
glimpse(RATSL)
summary(RATSL)

p1 <- ggplot(RATSL, aes(x = Time, y = stdWeight, group = ID))
p2 <- p1 + geom_line(aes(linetype = Group, color = Group))
p3 <- p2 +  scale_x_continuous(name = "Time (days)", breaks = seq(1, 64, 7)) +
  scale_y_continuous(name = "Standardized weight")
p4 <- p3 + theme_light() + ggtitle(label = "Figure 2 Repeated measures of weight (standardized")
p4
```

<br/>

```{r 6_fig3, fig.height=5, fig.width=8}
# Number of measurement points, baseline (week 0) included:
n <- RATSL$Time %>% unique() %>% length()

# Make a summary data:
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean=mean(Weight), se=sd(Weight)/sqrt(n) ) %>%
  ungroup()
glimpse(RATSS)

p1 <- ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group, color = Group))
p2 <- p1 + geom_line()+ scale_linetype_manual(values = c(1,2,3))
p3 <- p2 + geom_point(size=3) + scale_shape_manual(values = c(1,2,3))
p4 <- p3 + geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3)
p5 <- p4 + theme_light() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p6 <- p5 + scale_y_continuous(name = "mean(Weight) +/- se(Weight)") +
  ggtitle(label = "Figure 3 Mean weight of treatment groups w/ standard error bars")
p6
```

<br/>

The differences between the groups seem to be outside the standard error in all time points.

<br/>

```{r 6_fig4, fig.height=5, fig.width=8}
p1 <- ggplot(RATSL, aes(x = factor(Time), y = Weight, fill = Group))
p2 <- p1 + geom_boxplot(position = position_dodge(width = 0.9))
p3 <- p2 + theme_light() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p4 <- p3 + scale_x_discrete(name = "Time") + ggtitle(label = "Figure 4 Treatment group weight distributions in time")
p4
```

<br/>

Boxplot also shows clear differences between the groups. 

<br/>

```{r 6_fig5, fig.height=5, fig.width=8}
# Make a summary data of the post treatment weeks
RATSSS <- RATSL %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()
glimpse(RATSSS)

p1 <- ggplot(RATSSS, aes(x = Group, y = mean))
p2 <- p1 + geom_boxplot(aes(fill = Group))
p3 <- p2 + theme_light() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p4 <- p3 + stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white")
p5 <- p4 + scale_y_continuous(name = "mean(Weight) after time 1") +
  ggtitle(label = "Figure 4 Mean weight per treatment gourp after time 1")
p5
```

<br/>

The differences are also quite clear when examining the means of weight measurements per a rat after first weighing. However, there seems to be an outlier in group 2.

<br/>

```{r 6_outliers}
# Remove the outliers:
RATSSS1 <- RATSSS %>%
  filter(mean < 550)
glimpse(RATSSS1)

p1 <- ggplot(RATSSS1, aes(x = Group, y = mean))
p2 <- p1 + geom_boxplot(aes(fill = Group))
p3 <- p2 + theme_light() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p4 <- p3 + stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white")
p5 <- p4 + scale_y_continuous(name = "mean(Weight) after time 1") +
  ggtitle(label = "Figure 4 Mean weight per treatment gourp after time 1")
p5
```

<br/>

After removing the outlier, the differences seem even more clear.

<br/>

### ANOVA

```{r 6_ANOVA}
fit1 <- lm(mean ~ Group, data = RATSSS1)
summary(fit1)

# same with ANOVA to get Tuckey multiple pairwise comparison
res.aov <- aov(mean ~ Group, data = RATSSS1)
summary(res.aov)
TukeyHSD(res.aov)

RATSSS1 %>% group_by(Group) %>% summarise(mean = mean(mean)) %>%
  kable(digits = 2, caption = "<b> Table 3</b> Group means") %>%
  kable_styling(bootstrap_options = c("hover", "striped"), full_width = F) %>%
  column_spec(1, bold = T, width = "10em") %>% 
  column_spec(2, width = "10em")
```

<br/>

It seems the groups are significantly different (F(2) = 501.8, p < .001). Pairwise comparison with Tukey HSD shows that all groups differ significantly from each other.  However, caution is advised in inference of these results due to the small sample size. Group means can be found in Table 3.

<br/>

###

```{r 6_ANOVA2}
# Add the baseline from the original data as a new variable to the summary data:
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", header = T)
baseline <- RATS$WD1
RATSSS2 <- RATSSS %>%
  mutate(baseline)
# Fit the ANCOVA model and see the results:
fit2 <- lm(mean ~ baseline + Group, data = RATSSS2)
summary(fit2)
anova(fit2)
```

<br/>

It seems that adding the baseline to the model removes the significant effect of the treatment group. In other words, differences between the groups can be explained with the differences in the rats' weight in the beginning of the experiment.

<br/><br/>

## Task 2: The BPRS

Implement the analyses of Chapter 9 of MABS using the BPRS data. (0-8 points: 0-4 points for graphs or analysis results + 0-4 points for their interpretations)

### The BPRSL Data

The BPRS dataset has 360 observations and 5 variables from 40 subjects: treatment vs. no treatment group, subject id, and weekly measure of breaf psychiatric rating scale (BPRS) from week 0 (baseline) to week 8.


```{r 6_data2}
BPRSL <- read.table("data/BPRSL.csv", sep = ",", header = T, row.names = 1)
BPRSL <- within(BPRSL, {
  subject <- factor(subject)
  treatment <- factor(treatment)
})
glimpse(BPRSL)
head(BPRSL)
tail(BPRSL)
```

```{r, echo=TRUE}
p1 <- ggplot(BPRSL, aes(x = week, y = bprs, group = subject))
p2 <- p1 + geom_text(aes(label = treatment, color = treatment))
p3 <- p2 + scale_x_continuous(name = "BPRS", breaks = seq(0, 8, 1))
p4 <- p3 + scale_y_continuous(name = "Weight (grams)")
p5 <- p4 + theme_light()
p6 <- p5 + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p6
```

<br/>

### Linear Mixed Effect Models

```{r 6_reg}
BPRS_reg <- lm(bprs ~ week + treatment, data = BPRSL)
summary(BPRS_reg)

```

<br/>

As we can see, there is a significant effect of time (week) but the effect of treatment group is not significant on subjects psychiatric rating (bprs).

<br/>

```{r 6_fig6, fig.height=5, fig.width=8}
# Number of weeks, baseline (week 0) included
n <- BPRSL$week %>% unique() %>% length()

# Summary data with mean and standard error of bprs by treatment and week 
BPRSS <- BPRSL %>%
  group_by(treatment, week) %>%
  summarise( mean = mean(bprs), se = sd(bprs)/sqrt(n)  ) %>%
  ungroup()

# Plot the mean profiles
ggplot(BPRSS, aes(x = week, y = mean, linetype = treatment, shape = treatment, color = treatment)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme_light() +
  theme(legend.position = c(0.8,0.8)) +
  scale_y_continuous(name = "mean(bprs) +/- se(bprs)")


```

<br/>

```{r, echo=TRUE, fig.width=10, fig.height=10}
BPRS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt", header = T)
pairs(BPRS[,c(-1,-2)], cex = 0.7)
```

<br/>

```{r 6_ref}
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref)
```

```{r 6_ref1}
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref1)
anova(BPRS_ref1, BPRS_ref)
```

```{r 6_ref2}
BPRS_ref2 <- lmer(bprs ~ week * treatment + (week | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref2)
# dummies (in Table) vs summary output: D1 = Group2, D2 = Group3
anova(BPRS_ref1, BPRS_ref2)
```

<br>

Comparing the random effect models with ANOVA shows that model ref1 fits the data better than ref (Chisq = 7.27, p = .026). However ref2 does not improve the fit statistically significantly in comparison to ref1 (Chisq = 3.17, p = .075).

<br/>


